---
title: "Exploratory Data Analysis of Credit Risk Dataset"
author: "Aleksei Prishchepo"
date: "2025-09-15"
fig-format: svg
number-sections: true
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    theme: cosmo
    code-fold: true
    code-summary: "Click to show/hide code"
    highlight-style: tango

    
  # pdf:
  #   papersize: letter
  #   margin-left: 1.5cm
  #   margin-right: 1.5cm
  #   margin-top: 1.5cm
  #   margin-bottom: 2.5cm
  #   colorlinks: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.width = 8)
library(tidyverse)
library(janitor)
library(skimr)
library(GGally)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
```

## Load Data

```{r load_data}
data <- readxl::read_excel("Credit_Risk_Dataset_Onyx_Data_September_25.xlsx")
```

## Basic Descriptives

```{r skim_data}
data |> skim()
```

```{r glimpse_data}
data |> glimpse()
```

Using latitude and longitude variables for predictions doesn't make much
sense, so we should remove them, along with `client_ID`, which serves
only as an identifier.

```{r}
data <- data |>
  select(-city_latitude, -city_longitude, -client_ID)
```

## Univariate Distributions


```{r }
#| fig-height: 6
#| fig-width: 6
# Numeric distributions
data |>
  select(where(is.numeric)) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = value)) +
  facet_wrap(~name, scales = "free", ncol = 3) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") + 
  theme(axis.title = element_blank())
```

## Correlations

```{r fig-corrplot}
library(corrplot)

cor_data <- data |>
  select_if(is.numeric) |>
  cor(use = "pairwise.complete.obs")

cor_data |>
  corrplot(
    type = "upper", order = "hclust",
    tl.col = "black", tl.srt = 45
  )
```

There are three pairs of highly correlated variables. Let's check them
and drop redundant ones.

```{r}
cor(data$loan_percent_income, data$loan_to_income_ratio, use = "pairwise.complete.obs")
```

```{r}
cor(data$person_age, data$cb_person_cred_hist_length, use = "pairwise.complete.obs")
```

```{r}
cor(data$person_income, data$other_debt, use = "pairwise.complete.obs")
```

Obviously, `loan_percent_income` is in fact `loan_to_income_ratio`, so
let's drop it. Other two correlations are not perfect, so we keep them.

```{r}
data <- data |>
  select(-loan_percent_income)
```

## Missing Values

```{r fig-missing-upset}
#| fig-height: 3
#| fig-width: 4
library(naniar)

data |>
  select(person_emp_length, loan_int_rate) |>
  gg_miss_upset(
    sets.bar.color = "#56B4E9", point.size = 3,
    main.bar.color = "grey39", matrix.color = "grey39"
  )
```

There is no obvious pattern in missingness. We're going to impute missing values using median.

```{r impute_missing}
data <- data |>
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
```

## Outliers

```{r fig-outliers}
# | fig-height: 6
# | fig-width: 6

data |>
  select(where(is.numeric)) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = name, y = log(value))) +
  geom_boxplot(fill = "steelblue", outlier.color = "red", outlier.size = 1) +
  theme(axis.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

```

There are outliers in `person_age` and `person_emp_length` that seem to be data errors. There are also extreme values in `other_debt` and `person_income`. We'll zoom in to see them better.


```{r fig-outliers-zoom}
require(gridExtra)
library(scales)

plot1 <- data |>
  select(where(is.numeric)) |>
  select(person_age, person_emp_length) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = name, y = value)) +
  geom_boxplot(fill = "steelblue", outlier.color = "red", outlier.size = 1) +
  theme(axis.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

plot2 <-  data |>
  select(where(is.numeric)) |>
  select(other_debt, person_income) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = name, y = log10(value))) +
  geom_boxplot(fill = "steelblue", outlier.color = "red", outlier.size = 1) +
  scale_y_continuous(labels = trans_format(function(x) round(10^x, -1), label_number_auto())) +
  theme(axis.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(plot1, plot2, ncol=2)

```
Let's correct `person_age` and `person_emp_length` by replacing outliers with median. Other utliers seem to be valid values. We're not going to remove them.

```{r correct_outliers}
data <- data |>
  mutate(
    person_age = ifelse(person_age > 100,
      median(person_age, na.rm = TRUE),
      person_age
    ),
    person_emp_length = ifelse(person_emp_length > 70,
      median(person_emp_length, na.rm = TRUE),
      person_emp_length
    )
  )
```

## Selecting Predictors

We're going to build a logistic regression fit to predict loan default
(`loan_status` variable). 0 means no default, 1 means default.

Let's convert categorical variables to factors.

```{r convert_factors}
data <- data |>
  mutate(across(where(is.character), as.factor))
```

```{r linear-fit}
data <- data |> mutate(loan_status = (loan_status == 1))

data1 <- data |> select(-city, -state)

fit <- glm(loan_status ~ ., family = binomial(link = "logit"), x = TRUE, data = data1)
summary(fit)
```
## Moidel Diagnostics

```{r}
anova(fit, test="Chisq")
```

```{r}
library(performance)
r2(fit)
```

Check for overdispersion:

```{r}
dispersion <- summary(fit)$deviance / summary(fit)$df.residual
dispersion
```
Dispersion is less than  1, so no overdispersion.

## Marginal Effects


```{r}
library(erer)

ma <- maBina(fit)$out

ma <- ma |>
  rownames_to_column("variable")
ma
```
